Prognostic and biological relevance of collagen-related genes in prostate cancer

### 2025-02-25 paper/markdown/manuscript_figures_tables.Rmd
ROC analyses for BCR
Analysis of prediction of VCR at one, two. three, and five year follow-up by ROC. Inference for the GBM models

Machine learning models of BCR-free survival were constructed for the following sets of explanatory variables: 
(1) first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, 
(2) pathological tumor stage, ISUP grade, and ComBat-adjusted $log_2$ mRNA levels of the PSA-coding gene _KLK3_ [@Lunger2021; @Boyukozer2020] as clinical predictors of BCR, and 
(3) pathological tumor stage, ISUP grade, ComBat-adjusted $log_2$ _KLK3_ mRNA levels, as well as the first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts. 
The models were trained in the pooled GEO cohort with six algorithms (RIDGE regularized Cox proportional hazard regression, Elastic Net regularized Cox proportional hazard regression, LASSO regularized Cox proportional hazard regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]) [@Simon2011; @Tibshirani1996; @Zou2005; @Friedman2010; @Fouodo2018; @Weston1998; @Breiman2001; @Ishwaran2008; @Friedman2001; @Greenwell2022; @Friedman2002]. 
Performance of the models was evaluated by comparison of the predicted and actual BCR-free survival in the pooled GEO training cohort, and TCGA and DKFZ validation collectives with Harrell's concordance index and Integrated Brier Scores (__Supplementary Tables S7__ - __S8__). 
Additionally, BCR-free survival in patients stratified by tertiles of predictor scores returned by the machine learning models was explored with Kaplan-Meier plots and Peto-Peto tests. 
Finally, confidence and calibration of the models was assessed by plots of Brier Scores for unique follow-up times.

@@

Overall survival of PCA patients was modeled in the GSE16560 watchful waiting cohort [@Sboner2010]. 
First- and second-order terms of ComBat-adjusted $log_2$-transformed expression levels of the collagen-related genes served as explanatory factors. 
Machine learning models using the algorithms outlined above for the BCR-free survival were trained in a training subset of the GSE16560 cohort, and validated in the test subset. 
Performance of the models of overall survival was evaluated in the training and test subsets of the GSE16560 cohort as described above for BCR-free survival (__Supplementary Tables S9__ - __S10__).

>>>>

Machine learning models of BCR-free survival were constructed for the following sets of explanatory variables: 
(1) first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, 
(2) pathological tumor stage, ISUP grade, and ComBat-adjusted $log_2$ mRNA levels of the PSA-coding gene _KLK3_ [@Lunger2021; @Boyukozer2020] as clinical predictors of BCR, and 
(3) pathological tumor stage, ISUP grade, ComBat-adjusted $log_2$ _KLK3_ mRNA levels, as well as the first- and second-order terms of ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts. 
The models were trained in the pooled GEO cohort with six algorithms (RIDGE regularized Cox proportional hazard regression, Elastic Net regularized Cox proportional hazard regression, LASSO regularized Cox proportional hazard regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]) [@Simon2011; @Tibshirani1996; @Zou2005; @Friedman2010; @Fouodo2018; @Weston1998; @Breiman2001; @Ishwaran2008; @Friedman2001; @Greenwell2022; @Friedman2002]. 
Performance of the models was evaluated by comparison of the predicted and actual BCR-free survival in the pooled GEO training cohort, and TCGA and DKFZ validation collectives with Harrell's concordance index and Integrated Brier Scores (__Supplementary Tables S7__ - __S8__). 
BCR-free survival in patients stratified by tertiles of predictor scores returned by the machine learning models was explored with Kaplan-Meier plots and Peto-Peto tests. 
Additionally, confidence and calibration of the models was assessed by plots of Brier Scores for unique follow-up times. 
Finally, ability to predict BCR at one, two, three, and five years after diagnosis was assessed by receiver-operating characteristic (ROC) [@Heagerty2000].

@@

Overall survival of PCA patients was modeled in the GSE16560 watchful waiting cohort [@Sboner2010]. 
First- and second-order terms of ComBat-adjusted $log_2$-transformed expression levels of the collagen-related genes served as explanatory factors. 
Machine learning models using the algorithms outlined above for the BCR-free survival were trained in a training subset of the GSE16560 cohort, and validated in the test subset. 
Performance of the models of overall survival was evaluated in the training and test subsets of the GSE16560 cohort as described above for BCR-free survival (__Supplementary Tables S10__ - __S11__).

### 2025-02-25 paper/markdown/manuscript_supplement.Rmd
ROC analyses for BCR
Analysis of prediction of VCR at one, two. three, and five year follow-up by ROC. Inference for the GBM models

Multi-parameter modeling of biochemical relapse (BCR) free survival and overall survival was done with the packages 
_glmnet_ [@Friedman2010] and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) (Ridge, Elastic Net and LASSO cox regression), 
_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
_survival_ [@Therneau2000], 
_surviminer_ [@Kassambara2016],  
[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), and 
_caret_ [@Kuhn2008]. 
Univariable analysis of relapse-free survival was accomplished with packages [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer'), _survival_, _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).

@@

For the pooled GEO training cohort, and the TCGA and DKFZ validation collectives, predictor scores were returned by calling `predict()` method for the modeling objects. 
Next, univariable Cox proportional hazard regression models were constructed with the BCR-free survival as the response and each of the predictor scores as the sole explanatory variable (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
The proportional hazard assumption of the Cox proportional hazard models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
The Cox proportional hazard models were used for evaluation of machine learning accuracy with Harrell's concordance index [@Harrell1996] and assessment of global calibration with Integrated Brier Score [@Graf1999]. 
Those performance statistics were computed by calling `summary(type = 'fit')` method from [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) package for the Cox proportional hazard modeling objects (__Supplementary Table S\@ref(tab:tab-bcr-eval)__). 
To assess calibration and confidence of BCR risk prediction of the machine learning models, we plotted Brier Scores defined as a mean square difference between the predicted BCR probability and the 0/1-coded observed BCR event [@Brier1950; @Graf1999] for unique survival time points. 
With this graphical method we evaluated predictions made by the best-performing GBM models (__Supplementary Figure S\@ref(fig:fig-clinic)B__). 
In addition, ability of the machine learning models to assign PCA patients to low-, intermediate-, and high-risk groups was investigated by comparing BCR-free survival between patients stratified by tertiles of the predictor scores with Kaplan-Meier plots and Peto-Peto test (functions `calibrate.coxex()`, `plot.calibrator()`, and `surv_pvalue()`, packages [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) and _survminer_) as presented for the GBM models in __Figure 2B__ and __Supplementary Figure \@ref(fig:fig-cl-plot)__.

@@

@@

__Figure \@ref(fig:fig-gbm-cohort). Investigation of confounding study effect on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__

@@

__Figure \@ref(fig:fig-clinic). Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.__

@@

```{r fig-cl-plot, fig.width = figur::convert(suppl_fig$cl_plot, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_plot, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.'}

@@

__Figure \@ref(fig:fig-cl-plot). GBM modeling of BCR-free survival in PCA with clinical prediction and expression of collagen-related transcripts: survival in tertiles of predictor scores.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-cl-plot)._ 
_BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots. Numbers of patients and BCR cases in the predictor score tertiles are indicated in the figure legend. P values for differences in survival between the tertiles obtained by Peto-Peto test are displayed in the plots. Note improved resolution of the survival curves for the combined clinical/expression model as compared with the clinical-only model in the pooled GEO and DKFZ cohort._

@@

__Figure \@ref(fig:fig-os). Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.__

>>>>

Multi-parameter modeling of biochemical relapse (BCR) free survival and overall survival was done with the packages 
_glmnet_ [@Friedman2010] and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) (Ridge, Elastic Net and LASSO cox regression), 
_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
_survival_ [@Therneau2000], 
_surviminer_ [@Kassambara2016],  
[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), 
_caret_ [@Kuhn2008], and 
_survivalROC_ [@Heagerty2000; @Heagerty2022].

Univariable analysis of relapse-free survival was accomplished with packages [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer'), _survival_, _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).

@@

For the pooled GEO training cohort, and the TCGA and DKFZ validation collectives, predictor scores were returned by calling `predict()` method for the modeling objects. 
Next, univariable Cox proportional hazard regression models were constructed with the BCR-free survival as the response and each of the predictor scores as the sole explanatory variable (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
The proportional hazard assumption of the Cox proportional hazard models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
The Cox proportional hazard models were used for evaluation of machine learning accuracy with Harrell's concordance index [@Harrell1996] and assessment of global calibration with Integrated Brier Score [@Graf1999]. 
Those performance statistics were computed by calling `summary(type = 'fit')` method from [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) package for the Cox proportional hazard modeling objects (__Supplementary Table S\@ref(tab:tab-bcr-eval)__). 
To assess calibration and confidence of BCR risk prediction of the machine learning models, we plotted Brier Scores defined as a mean square difference between the predicted BCR probability and the 0/1-coded observed BCR event [@Brier1950; @Graf1999] for unique survival time points. 
With this graphical method we evaluated predictions made by the best-performing GBM models (__Supplementary Figure S\@ref(fig:fig-clinic)B__). 
Ability of the machine learning models to assign PCA patients to low-, intermediate-, and high-risk groups was investigated by comparing BCR-free survival between patients stratified by tertiles of the predictor scores with Kaplan-Meier plots and Peto-Peto test (functions `calibrate.coxex()`, `plot.calibrator()`, and `surv_pvalue()`, packages [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) and _survminer_) as presented for the GBM models in __Figure 2B__ and __Supplementary Figure \@ref(fig:fig-cl-plot)__. 
Prediction of BCR at one, two, three, and five years after diagnosis by the GBM models was assessed by receiver-operating characteristic (ROC) with area under the curve (AUC) as a metric of prediction accuracy, sensitivity, and specificity. 
The ROC analysis was performed for predictor scores of the GBM models as marker variables with the nearest-neighbor algorithm proposed by Heagerty et al. [@Heagerty2000; @Heagerty2022] (function `survivalROC()`, package _survivalROC_). 
The ROC AUC values are outlined in __Supplementary Table S\@ref(tab:tab-roc)__.

@@

```{r tab-roc, tab.cap = "Prediction of biochemical relapse (BCR) at one, two, three, and five year after diagnosis by gradient boosted models employing clinical predictors and expression of the collagen-related transcripts as explanatory factors."}

flextable::flextable(suppl_tab$roc) %>% 
  merge_v(1:2) %>% 
  set_widths(c(2.5, 2.3, 4, rep(2.5, 3))) %>% 
  footnote(1, 3, 
            value = as_paragraph('GBM, expression, explanatory factors: expression levels of the collagen-related transcripts; GBM, clinic, explanatory factors: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue; GBM, clinic + expression: pathological tumor stage, ISUP grade, and log2 expression of the PSA coding gene KLK3 in the tumor tissue, log2 expression of the collagen-related transcripts.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6, 
           value = as_paragraph('ROC AUC: area under the curve of receiver-operating characteristic.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

__Supplementary Figure S\@ref(fig:fig-gbm-cohort). Investigation of confounding study effect on prediction of BCR-free survival by the GBM model in the pooled GEO cohort.__

@@

__Supplementary Figure S\@ref(fig:fig-clinic). Modeling of BCR-free survival by GBM algorithm with clinical predictors and mRNA expression of the collagen-related genes.__

@@

```{r fig-cl-plot, fig.width = figur::convert(suppl_fig$cl_plot, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_plot, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: survival in tertiles of predictor scores.'}

@@

__Supplementary Figure S\@ref(fig:fig-cl-plot). GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: survival in tertiles of predictor scores.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._

_BCR-free survival in tertiles of the predictor scores of the clinical-only and the combined clinical - expression GBM models was visualized with Kaplan-Meier plots. Numbers of patients and BCR cases in the predictor score tertiles are indicated in the figure legend. P values for differences in survival between the tertiles obtained by Peto-Peto test are displayed in the plots. Note improved resolution of the survival curves for the combined clinical/expression model as compared with the clinical-only model in the pooled GEO and DKFZ cohort._

```{r fig-cl-iference, fig.width = figur::convert(suppl_fig$cl_inference, to = 'in')$w, fig.height = figur::convert(suppl_fig$cl_inference, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: inference.'}

suppl_fig$cl_inference$plot

__Supplementary Figure S\@ref(fig:fig-cl-iference). GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: inference.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._

_Predictor scores of the GBM models were computed for the training collective, and the TCGA and DKFZ validation cohorts. Subsequently, univariable Cox proportional hazard models of BCR-free survival were constructed with normalized values of the predictor scores (Z-scores) as explanatory factors. Inference statistics of those Cox proportional hazard models, hazard ratios (HR) with 95% confidence intervals (95%CI),m are presented in a Forest plot._

```{r fig-roc, fig.width = figur::convert(suppl_fig$roc, to = 'in')$w, fig.height = figur::convert(suppl_fig$roc, to = 'in')$h, fig.cap = 'GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: prediction of BCR.'}

suppl_fig$roc$plot

__Supplementary Figure S\@ref(fig:fig-roc). GBM modeling of BCR-free survival in PCA with clinical predictors and expression of collagen-related transcripts: prediction of BCR.__

_Three Gradient Boosted Machines (GBM) models of biochemical relapse (BCR) free survival were trained in the pooled GEO cohort: a model with clinical predictors (age, pathological tumor stage, ISUP grade, and expression of KLK3 transcript coding for PSA protein), a model with ComBat-adjusted $log_2$ expression levels of the collagen-related transcripts, and a model with the clinical and transcriptomic predictors, as presented in Supplementary Figure S\@ref(fig:fig-clinic)._

_Ability of the GBM models to predict BCR at one, two, three, and five years after diagnosis in the pooled GEO training cohort, and the TCGA and DKFZ validation collectives was investigated by receiver-operating characteristic (ROC) with are under the curve (AUC) as a metric of accuracy, sensitivity, and specificity._

_(A) Values of AUC or prediction of BCR at one, two, three, and five years after diagnosis presented in a dot plot. Each point represents a single time point. Values obtained with the same survival model are connected with lines._

_(B) ROC curves for prediction of BCR at one, two, three, and five years after diagnosis (TR: true positive, FP: false positive). color of the curve represents the survival model. AUC values are indicated in the plots. Total numbers of patients in the modeling data and numbers of BCR cases at the time point are displayed in the plot captions._

@@

__Supplementary Figure S\@ref(fig:fig-os). Modeling of overall survival in PCA with expression levels of the collagen-related transcripts as explanatory factors.__
